\section{Notation, definitions, and basic information-theoretic facts}
\label{sec:prelim}

We collect the notation, the formal setup, and the standard
information-theoretic facts that the rest of the proof invokes.

\paragraph{Notation.}
We work over a fixed finite universe $[n] := \{1, 2, \dots, n\}$ for some $n \ge 1$.
Subsets of $[n]$ are identified with their indicator vectors in $\{0,1\}^n$
via $A \leftrightarrow (A_1, \dots, A_n)$ where $A_i = \1\{i \in A\}$.
We write $\log$ for the base-$2$ logarithm throughout, and adopt the
convention $0 \log 0 = 0$.
For a discrete random variable $X$ taking values in a finite set $\mathcal{X}$,
$\H(X) := -\sum_{x \in \mathcal{X}} \Pr[X=x] \log \Pr[X=x]$ is its (Shannon) entropy,
and $\H(X \mid Y) := \E_Y[\H(X \mid Y = y)]$ is its conditional entropy.
Mutual information is $\Iinfo(X;Y) := \H(X) - \H(X \mid Y)$.
For two Bernoulli random variables $X, X' \in \{0,1\}$,
$X \cup X' := \max(X, X')$, which is also the indicator of the event that at
least one of $X, X'$ equals $1$.
For two subsets $A, B \subseteq [n]$, $A \cup B$ is the usual set union,
and its indicator vector is $((A \cup B)_1, \dots, (A \cup B)_n)$ where
$(A \cup B)_i = A_i \cup B_i = \max(A_i, B_i)$.
Throughout, $c, C, C_1, \dots$ denote universal positive constants whose
values may change from line to line.

\paragraph{Binary entropy.}
The \emph{binary entropy function} is
\begin{align*}
\hbin(p) \;:=\; -p \log p - (1-p) \log (1-p), \qquad p \in [0,1],
\end{align*}
with $\hbin(0) := 0 =: \hbin(1)$. We will use repeatedly that $\hbin$ is
continuous, $\hbin(0) = \hbin(1) = 0$, $\hbin(1/2) = 1$, $\hbin$ is concave
on $[0,1]$, and for $X \sim \Bern(p)$, $\H(X) = \hbin(p)$.

\begin{definition}[Union-closed family]\label{def:union-closed}
A family $\Fcal \subseteq 2^{[n]}$ of subsets of $[n]$ is \emph{union-closed}
if for every $A, B \in \Fcal$, the set $A \cup B$ also belongs to $\Fcal$.
\end{definition}

\begin{remark}\label{rem:nontriviality}
We focus on union-closed families $\Fcal$ with $|\Fcal| \ge 2$; the case
$\Fcal = \{\varnothing\}$ is degenerate (the universe of an empty set has no
element to be ``abundant'' in), and the case $\Fcal = \varnothing$ is vacuous.
Frankl's original formulation~\cite{frankl1995extremal} required only
$\Fcal \ne \{\varnothing\}$; our slightly stronger $|\Fcal| \ge 2$ avoids the
edge case while changing nothing substantive (a union-closed family with
$|\Fcal| = 1$ contains a single set $A$, and if $A = \varnothing$ then there is
no element to abundance-test; if $A \ne \varnothing$ then any $i \in A$ is in
$100\% \ge 1\%$ of the sets, so the bound holds trivially).
\end{remark}

The following four facts about Shannon entropy are standard and stated
without proof; we cite Cover and Thomas~\cite{cover2006elements}.

\begin{fact}[Chain rule, {\cite[Theorem 2.5.1]{cover2006elements}}]\label{fac:chain-rule}
For discrete random variables $X_1, \dots, X_n$ taking values in finite sets,
\begin{align*}
\H(X_1, \dots, X_n) \;=\; \sum_{i=1}^{n} \H(X_i \mid X_1, \dots, X_{i-1}),
\end{align*}
with the convention that $\H(X_1 \mid X_{<1}) = \H(X_1)$.
\end{fact}

\begin{fact}[Data-processing for entropy, {\cite[Theorem 2.8.1]{cover2006elements}}]\label{fac:data-processing}
Let $X, Y$ be discrete random variables and let $f$ be any (measurable)
function defined on the range of $Y$. Then
\begin{align*}
\H(X \mid Y) \;\le\; \H(X \mid f(Y)).
\end{align*}
\end{fact}

\begin{fact}[Uniform maximizes entropy, {\cite[Theorem 2.6.4]{cover2006elements}}]\label{fac:uniform-max}
For a discrete random variable $X$ supported on a finite set $\mathcal{X}$,
$\H(X) \le \log |\mathcal{X}|$, with equality if and only if $X$ is uniform on $\mathcal{X}$.
\end{fact}

\begin{fact}[Concavity of binary entropy]\label{fac:hbin-concave}
The binary entropy function $\hbin : [0,1] \to [0, 1]$ is concave.
Equivalently, for any $p, p' \in [0,1]$ and $\lambda \in [0,1]$,
$\hbin(\lambda p + (1-\lambda) p') \ge \lambda \hbin(p) + (1-\lambda) \hbin(p')$.
\end{fact}

\begin{remark}\label{rem:hbin-concave}
\Cref{fac:hbin-concave} is immediate from $\hbin''(p) = -\frac{1}{(\ln 2)\, p (1-p)} < 0$ for $p \in (0,1)$,
together with the boundary values $\hbin(0) = \hbin(1) = 0$.
\end{remark}

\paragraph{Bernoulli unions.}
We will use the following identity throughout without further comment:
if $X \sim \Bern(p)$ and $X' \sim \Bern(p')$ are independent, then
$X \cup X' = \max(X, X') \sim \Bern(p + p' - p p')$, since
$\Pr[\max(X, X') = 0] = \Pr[X=0]\Pr[X'=0] = (1-p)(1-p')$.
In particular, $\H(X \cup X') = \hbin(p + p' - p p')$.
