\section{The information-theoretic strengthening}
\label{sec:theorem1}

The next theorem is the information-theoretic strengthening that drives the
combinatorial conclusion in \Cref{sec:main}. Its proof reveals the bits of
$A$ and $A \cup B$ one at a time, applies \Cref{lem:gap} to each conditional
slice, and aggregates by the chain rule.

\begin{theorem}[Information-theoretic strengthening]\label{thm:entropy-gap}
Let $A$ and $B$ be independent samples from a distribution over subsets of
$[n]$. Assume that for every $i \in [n]$, $\Pr[i \in A] \le 0.01$. Then
\begin{align}\label{eq:thm-entropy-gap}
\H(A \cup B) \;\ge\; 1.26 \, \H(A).
\end{align}
In particular, if $\H(A) > 0$, then $\H(A \cup B) > \H(A)$.
\end{theorem}

\begin{proof}
The proof has two steps: (i) prove a per-coordinate version
$\H((A \cup B)_i \mid (A \cup B)_{<i}) \ge 1.26 \, \H(A_i \mid A_{<i})$ for
each $i \in [n]$, and (ii) sum over $i$ by the chain rule
(\Cref{fac:chain-rule}).

\paragraph{Step 1: A useful function-of-history identity.}
Identify $A, B \subseteq [n]$ with their indicator vectors
$(A_1, \dots, A_n), (B_1, \dots, B_n) \in \{0, 1\}^n$. Write
$A_{<i} := (A_1, \dots, A_{i-1})$ (an empty tuple when $i = 1$) and similarly
$B_{<i}$. Observe that
\begin{align}\label{eq:func-of-hist}
(A \cup B)_{<i} \;=\; \bigl( A_1 \cup B_1, \dots, A_{i-1} \cup B_{i-1} \bigr)
\;=\; \phi(A_{<i}, B_{<i}),
\end{align}
where $\phi : \{0,1\}^{i-1} \times \{0,1\}^{i-1} \to \{0,1\}^{i-1}$ is the
coordinatewise-maximum map $\phi(a, b) = (\max(a_j, b_j))_{j < i}$. So
$(A \cup B)_{<i}$ is a (deterministic) function of $(A_{<i}, B_{<i})$.

\paragraph{Step 2: Per-coordinate conditional-entropy bound.}
Fix any $i \in [n]$. We show
\begin{align}\label{eq:per-coord}
\H((A \cup B)_i \mid (A \cup B)_{<i}) \;\ge\; 1.26 \, \H(A_i \mid A_{<i}).
\end{align}
By Eq.~\eqref{eq:func-of-hist} and \Cref{fac:data-processing}
(data-processing for entropy, applied with $X = (A \cup B)_i$,
$Y = (A_{<i}, B_{<i})$, and $f = \phi$),
\begin{align}\label{eq:dpi-step}
\H((A \cup B)_i \mid (A \cup B)_{<i})
\;=\; \H\!\bigl((A \cup B)_i \mid \phi(A_{<i}, B_{<i})\bigr)
\;\ge\; \H\!\bigl((A \cup B)_i \mid A_{<i}, B_{<i}\bigr).
\end{align}

We now apply \Cref{lem:gap} with the substitution
\begin{align*}
C \;\leftarrow\; A_{<i}, \quad
C' \;\leftarrow\; B_{<i}, \quad
X \;\leftarrow\; A_i, \quad
X' \;\leftarrow\; B_i.
\end{align*}
We verify the hypotheses of \Cref{lem:gap} under this substitution:
\begin{itemize}
\item $A_{<i}$ takes values in the finite set $S := \{0,1\}^{i-1}$ (with $S$
  a singleton when $i = 1$);
\item conditional on $A_{<i} = a$, $A_i$ is Bernoulli with parameter
  $p_a := \Pr[A_i = 1 \mid A_{<i} = a] \in [0, 1]$;
\item $B_{<i}$ has the same distribution as $A_{<i}$ (since $B$ has the same
  distribution as $A$) and is independent of $A_{<i}$ (since $A$ and $B$ are
  independent samples);
\item conditional on $B_{<i} = a$, $B_i$ is Bernoulli with parameter $p_a$
  (same conditional distribution as $A_i \mid A_{<i} = a$);
\item $B_i$ is independent of $(A_{<i}, A_i)$ (again by independence of $A$
  and $B$);
\item the mean hypothesis: $\E[A_i] = \Pr[i \in A] \le 0.01$ by assumption.
\end{itemize}
All the hypotheses of \Cref{lem:gap} are met. Furthermore, $A_i \cup B_i = (A \cup B)_i$ by definition.
The conclusion of \Cref{lem:gap} therefore reads
\begin{align}\label{eq:apply-gap}
\H\!\bigl((A \cup B)_i \mid A_{<i}, B_{<i}\bigr)
\;=\; \H(A_i \cup B_i \mid A_{<i}, B_{<i})
\;\ge\; 1.26 \, \H(A_i \mid A_{<i}).
\end{align}
Combining Eq.~\eqref{eq:dpi-step} and Eq.~\eqref{eq:apply-gap} gives
Eq.~\eqref{eq:per-coord}.

\paragraph{Step 3: Aggregate via the chain rule.}
By the chain rule (\Cref{fac:chain-rule}) applied to the indicator vectors
of $A \cup B$ and of $A$, respectively,
\begin{align*}
\H(A \cup B)
\;=\; \H\!\bigl((A \cup B)_1, \dots, (A \cup B)_n\bigr)
\;=\; \sum_{i=1}^{n} \H\!\bigl((A \cup B)_i \mid (A \cup B)_{<i}\bigr),
\end{align*}
and
\begin{align*}
\H(A) \;=\; \sum_{i=1}^{n} \H(A_i \mid A_{<i}).
\end{align*}
Applying Eq.~\eqref{eq:per-coord} term-by-term,
\begin{align*}
\H(A \cup B) \;=\; \sum_{i=1}^{n} \H\!\bigl((A \cup B)_i \mid (A \cup B)_{<i}\bigr)
\;\ge\; \sum_{i=1}^{n} 1.26 \, \H(A_i \mid A_{<i}) \;=\; 1.26 \, \H(A).
\end{align*}
This is Eq.~\eqref{eq:thm-entropy-gap}.

\paragraph{Strict inequality when $\H(A) > 0$.}
If $\H(A) > 0$, then $1.26 \, \H(A) > \H(A)$, and the chain
$\H(A \cup B) \ge 1.26 \, \H(A) > \H(A)$ yields $\H(A \cup B) > \H(A)$
strictly. This completes the proof.
\end{proof}
