\section{Optimism via Backward Induction}\label{sec:optimism}

\begin{lemma}[Per-step Bellman gap on $\cE$]\label{lem:per-step}
On the event $\cE$, for every $k \in [K]$, $h \in [H]$, and $(s, a) \in \cS \times \cA$,
\begin{align}\label{eq:per_step_lower}
Q_h^k(s, a) - r_h(s, a) - [P_h V_{h+1}^k](s, a) &\;\le\; 2 \beta\, \norm{\phi(s, a)}_{(\Lam_h^k)^{-1}}, \\
Q_h^k(s, a) - r_h(s, a) - [P_h V_{h+1}^k](s, a) &\;\ge\; 0
\quad \text{or}\quad Q_h^k(s, a) = H. \label{eq:per_step_upper}
\end{align}
\end{lemma}

\begin{proof}
We work on $\cE$ throughout. Recall from Eq.~\eqref{eq:Q_def} that $Q_h^k(s, a) = \min\{ \widetilde{Q}_h^k(s, a),\, H \}$ with $\widetilde{Q}_h^k(s, a) \coloneqq \inner{\phi(s, a)}{\wh{w}_h^k} + \beta \norm{\phi(s, a)}_{(\Lam_h^k)^{-1}}$.

\textbf{Upper inequality (Eq.~\eqref{eq:per_step_lower}).} By Eq.~\eqref{eq:good_event} (which holds on $\cE$), $\inner{\phi(s,a)}{\wh{w}_h^k} \le r_h(s,a) + [P_h V_{h+1}^k](s,a) + \beta \norm{\phi(s,a)}_{(\Lam_h^k)^{-1}}$, hence
\begin{align*}
Q_h^k(s, a) \;\le\; \widetilde{Q}_h^k(s, a)
\;\le\; r_h(s, a) + [P_h V_{h+1}^k](s, a) + 2 \beta \norm{\phi(s, a)}_{(\Lam_h^k)^{-1}}.
\end{align*}

\textbf{Lower inequality (Eq.~\eqref{eq:per_step_upper}).} Consider two cases. If $Q_h^k(s,a) = H$, the second branch of the disjunction holds. Otherwise $Q_h^k(s, a) = \widetilde{Q}_h^k(s, a) \le H$, and by Eq.~\eqref{eq:good_event} applied to the lower direction,
\begin{align*}
Q_h^k(s, a)
\;=\; & ~ \inner{\phi(s, a)}{\wh{w}_h^k} + \beta\, \norm{\phi(s, a)}_{(\Lam_h^k)^{-1}} \\
\;\ge\; & ~ r_h(s, a) + [P_h V_{h+1}^k](s, a) - \beta \norm{\phi(s, a)}_{(\Lam_h^k)^{-1}} + \beta \norm{\phi(s, a)}_{(\Lam_h^k)^{-1}} \\
\;=\; & ~ r_h(s, a) + [P_h V_{h+1}^k](s, a),
\end{align*}
where the first step is the definition of $\widetilde{Q}_h^k$ in the un-truncated branch, the second uses Eq.~\eqref{eq:good_event} on $\cE$, and the third is algebra. This completes the proof.
\end{proof}

\begin{lemma}[Optimism]\label{lem:optimism}
On the event $\cE$, for every $k \in [K]$, every $h \in [H+1]$, and every $s \in \cS$,
\begin{align}\label{eq:optimism_main}
V_h^k(s) \;\ge\; V_h^*(s).
\end{align}
\end{lemma}

\begin{proof}
We proceed by backward induction on $h$, conditional on $\cE$.

\textbf{Base case ($h = H+1$).} Both $V_{H+1}^k$ and $V_{H+1}^*$ are identically zero by convention, so Eq.~\eqref{eq:optimism_main} holds with equality.

\textbf{Inductive step.} Assume Eq.~\eqref{eq:optimism_main} holds at step $h + 1$. Fix $s \in \cS$ and let $a^* = \pi_h^*(s) \in \argmax_a Q_h^*(s, a)$. Then for that $a^*$,
\begin{align*}
V_h^k(s)
\;\ge\; & ~ Q_h^k(s, a^*) \\
\;\ge\; & ~ \min\bigl\{r_h(s, a^*) + [P_h V_{h+1}^k](s, a^*),\; H\bigr\} \\
\;\ge\; & ~ \min\bigl\{r_h(s, a^*) + [P_h V_{h+1}^*](s, a^*),\; H\bigr\} \\
\;=\; & ~ \min\bigl\{Q_h^*(s, a^*),\; H\bigr\} \\
\;=\; & ~ V_h^*(s),
\end{align*}
where the first step is $V_h^k(s) = \max_a Q_h^k(s, a)$ by Eq.~\eqref{eq:V_def}, the second step is the lower inequality of \Cref{lem:per-step} (Eq.~\eqref{eq:per_step_upper}: either $Q_h^k(s, a^*) \ge r_h(s, a^*) + [P_h V_{h+1}^k](s, a^*)$ or $Q_h^k(s, a^*) = H$, both of which lower-bound by $\min\{r_h + P_h V_{h+1}^k, H\}$), the third step is monotonicity of $f \mapsto P_h f$ together with the inductive hypothesis $V_{h+1}^k \ge V_{h+1}^*$ and monotonicity of $\min\{\cdot, H\}$, the fourth is the Bellman equation $Q_h^*(s, a^*) = r_h(s, a^*) + [P_h V_{h+1}^*](s, a^*)$, and the last uses $Q_h^*(s, a^*) \le H$ (since $r \in [0, 1]$ and there are $H - h + 1 \le H$ remaining steps), so $\min\{Q_h^*(s, a^*), H\} = Q_h^*(s, a^*) = V_h^*(s)$. This closes the induction.
\end{proof}

\begin{remark}\label{rem:optimism_role}
\Cref{lem:optimism} converts the upper-bound bonus structure of LSVI-UCB into a one-sided guarantee: on $\cE$, $V_1^k(s_1^k) \ge V_1^*(s_1^k)$, so the per-episode regret $V_1^*(s_1^k) - V_1^{\pi^k}(s_1^k)$ is bounded above by the \emph{algorithm's} estimate gap $V_1^k(s_1^k) - V_1^{\pi^k}(s_1^k)$. This is what feeds into the regret decomposition in \Cref{lem:decomposition}.
\end{remark}
