\section{Regret Decomposition}\label{sec:decomposition}

This section reduces $\Reg(K)$ on the good event $\cE$ to three terms, $T_1, T_2, T_3$, that can be controlled by martingale concentration and the elliptical-potential lemma respectively.

\begin{lemma}[Decomposition into $T_1 + T_2 + T_3$]\label{lem:decomposition}
On the event $\cE$,
\begin{align}\label{eq:reg_decomposition}
\Reg(K) \;\le\; T_1 + T_2 + T_3,
\end{align}
where
\begin{align}
T_1 &\;\coloneqq\; \sum_{k=1}^K \sum_{h=1}^H \zeta_{h, 1}^k, \quad \text{with}\quad \zeta_{h, 1}^k \;\coloneqq\; [P_h (V_{h+1}^k - V_{h+1}^{\pi^k})](s_h^k, a_h^k) - (V_{h+1}^k - V_{h+1}^{\pi^k})(s_{h+1}^k), \label{eq:T1_def}\\
T_2 &\;\coloneqq\; \sum_{k=1}^K \sum_{h=1}^H 2 \beta\, \norm{\phi(s_h^k, a_h^k)}_{(\Lam_h^k)^{-1}}, \label{eq:T2_def}\\
T_3 &\;\coloneqq\; 0. \label{eq:T3_def}
\end{align}
(The placeholder $T_3 = 0$ is retained for symmetry with prior expositions where $T_3$ collects a vanishing term; here it is absorbed into $T_1$.)
\end{lemma}

\begin{proof}
On $\cE$, by \Cref{lem:optimism}, $V_1^*(s_1^k) \le V_1^k(s_1^k)$ for each $k$, so
\begin{align*}
\Reg(K) \;\le\; \sum_{k=1}^K \bigl[ V_1^k(s_1^k) - V_1^{\pi^k}(s_1^k) \bigr].
\end{align*}

\smallskip\noindent
\textbf{Telescoping the per-episode gap.} Fix episode $k$. For each $h \in [H]$ define the policy-following gap
\begin{align*}
\delta_h^k \;\coloneqq\; V_h^k(s_h^k) - V_h^{\pi^k}(s_h^k).
\end{align*}
At step $h$, the algorithm acts greedily wrt $Q_h^k$ so $V_h^k(s_h^k) = Q_h^k(s_h^k, a_h^k)$, and the policy-value definition gives $V_h^{\pi^k}(s_h^k) = Q_h^{\pi^k}(s_h^k, a_h^k) = r_h(s_h^k, a_h^k) + [P_h V_{h+1}^{\pi^k}](s_h^k, a_h^k)$. Using these and \Cref{lem:per-step}\,Eq.~\eqref{eq:per_step_lower},
\begin{align*}
\delta_h^k
\;=\; & ~ Q_h^k(s_h^k, a_h^k) - r_h(s_h^k, a_h^k) - [P_h V_{h+1}^{\pi^k}](s_h^k, a_h^k) \\
\;\le\; & ~ 2 \beta\, \norm{\phi(s_h^k, a_h^k)}_{(\Lam_h^k)^{-1}} + [P_h V_{h+1}^k](s_h^k, a_h^k) - [P_h V_{h+1}^{\pi^k}](s_h^k, a_h^k) \\
\;=\; & ~ 2 \beta\, \norm{\phi(s_h^k, a_h^k)}_{(\Lam_h^k)^{-1}} + [P_h (V_{h+1}^k - V_{h+1}^{\pi^k})](s_h^k, a_h^k) \\
\;=\; & ~ 2 \beta\, \norm{\phi(s_h^k, a_h^k)}_{(\Lam_h^k)^{-1}} + (V_{h+1}^k - V_{h+1}^{\pi^k})(s_{h+1}^k) + \zeta_{h, 1}^k,
\end{align*}
where the first step uses $V_h^k - V_h^{\pi^k} = Q_h^k - Q_h^{\pi^k}$ at $(s_h^k, a_h^k)$ and the Bellman expansion of $Q_h^{\pi^k}$, the second is \Cref{lem:per-step}\,Eq.~\eqref{eq:per_step_lower} applied to the Bellman residual of $V_{h+1}^k$ minus the (true) Bellman backup of $V_{h+1}^{\pi^k}$, the third step is linearity of $P_h$, and the last is the definition of $\zeta_{h, 1}^k$ in Eq.~\eqref{eq:T1_def}.

\smallskip\noindent
\textbf{Telescoping in $h$.} Setting $\delta_{H+1}^k = 0$ (since $V_{H+1}^k = V_{H+1}^{\pi^k} \equiv 0$) and summing the recursion $\delta_h^k \le 2 \beta \norm{\phi_h^k}_{(\Lam_h^k)^{-1}} + \delta_{h+1}^k + \zeta_{h, 1}^k$ over $h = 1, \ldots, H$ yields
\begin{align*}
\delta_1^k \;\le\; \sum_{h=1}^H \Bigl[ 2 \beta\, \norm{\phi(s_h^k, a_h^k)}_{(\Lam_h^k)^{-1}} + \zeta_{h, 1}^k \Bigr],
\end{align*}
where we used $\delta_{h+1}^k = (V_{h+1}^k - V_{h+1}^{\pi^k})(s_{h+1}^k)$ at the next step's realised state $s_{h+1}^k$.

\smallskip\noindent
\textbf{Summing over $k$.} Summing over $k$ and recognising the two terms from Eq.~\eqref{eq:T1_def} and Eq.~\eqref{eq:T2_def} gives $\Reg(K) \le T_1 + T_2$, which yields Eq.~\eqref{eq:reg_decomposition} on setting $T_3 = 0$. This completes the proof.
\end{proof}

\begin{remark}\label{rem:T1_T3_separation}
Some expositions (including \cite{jin2020provably}) separate the martingale term $T_1$ into two pieces: one for the transition-noise martingale and one for the bonus-self-normalisation martingale. We have merged them into $T_1$ because, on $\cE$, the self-normalisation has already been absorbed into the bound \Cref{lem:per-step} that produced $T_2$, leaving only the transition-noise martingale here. Hence $T_3 = 0$ in our formulation.
\end{remark}

\begin{lemma}[Martingale concentration for $T_1$]\label{lem:azuma_mds}
For any $\delta_0 \in (0, 1)$, with probability at least $1 - \delta_0$,
\begin{align}\label{eq:T1_bound}
|T_1| \;\le\; 4 H \sqrt{T \log(2 / \delta_0)}.
\end{align}
\end{lemma}

\begin{proof}
We claim $\{\zeta_{h, 1}^k\}$ is a martingale-difference sequence with $|\zeta_{h, 1}^k| \le 2 H$ a.s. Order the indices lexicographically: $(k, h) \preceq (k', h')$ iff $k < k'$, or $k = k'$ and $h \le h'$. Let $\cF_{h}^k$ denote the $\sigma$-algebra of data through the end of step $h$ of episode $k$.

The function $V_{h+1}^k$ is $\cF_{h-1}^k$-measurable (it depends only on data from episodes $\tau < k$, and on the algorithm's deterministic backward recursion), and so is $V_{h+1}^{\pi^k}$ given $\pi^k$ which is $\cF_0^k$-measurable. Hence $[P_h(V_{h+1}^k - V_{h+1}^{\pi^k})](s_h^k, a_h^k)$ is $\cF_{h-1}^k$-measurable, while $(V_{h+1}^k - V_{h+1}^{\pi^k})(s_{h+1}^k)$ has conditional expectation equal to $[P_h(V_{h+1}^k - V_{h+1}^{\pi^k})](s_h^k, a_h^k)$. Therefore $\E[\zeta_{h, 1}^k \mid \cF_{h-1}^k] = 0$.

Since $V_{h+1}^k, V_{h+1}^{\pi^k} \in [0, H]$, $|\zeta_{h, 1}^k| \le 2 H$ deterministically. By Azuma--Hoeffding (\cite{azuma1967weighted}), with probability at least $1 - \delta_0$,
\begin{align*}
|T_1| \;=\; \Bigl|\sum_{k, h} \zeta_{h, 1}^k\Bigr|
\;\le\; \sqrt{2 \cdot (2H)^2 \cdot T \cdot \log(2/\delta_0)}
\;=\; 2 \sqrt{2} \, H \sqrt{T \log(2/\delta_0)}
\;\le\; 4 H \sqrt{T \log(2/\delta_0)},
\end{align*}
giving Eq.~\eqref{eq:T1_bound}. This completes the proof.
\end{proof}
