\section{Preliminaries}\label{sec:prelim}

\paragraph{Notation.}
We write $[n]=\{1,\dots,n\}$. For a vector $v$, $\norm{v}$ is its Euclidean norm;
for a matrix $M$, $\opnorm{M}$ is the operator (spectral) norm and $\fnorm{M}$ the
Frobenius norm. For a symmetric matrix $M$, $\lmin(M)$ denotes its smallest eigenvalue.
Throughout, $c, C, C_1, C_2, \ldots$ denote universal positive constants whose values may
change from line to line and depend only on quantities specified in each statement's
hypothesis; constants that depend on a problem parameter carry it as a subscript
(e.g.\ $C_{\lzero}$). The training data are $\{(x_i, y_i)\}_{i=1}^n$ with inputs normalized
to $\norm{x_i} = 1$ and labels $\abs{y_i} \le 1$.

\paragraph{Model and training.}
We study the two-layer ReLU network
\begin{align}\label{eq:network}
f_{W}(x) \;=\; \frac{1}{\sqrt{m}} \sum_{r=1}^{m} a_r \, \sigma\!\left( w_r^{\top} x \right),
\qquad \sigma(z) = \max\{z, 0\},
\end{align}
with first-layer weights $W = (w_1, \dots, w_m) \in \R^{d \times m}$ and output weights
$a_r \in \{\pm 1\}$. At initialization $w_r(0) \sim \N(0, I_d)$ i.i.d.\ and
$a_r \sim \Unif\{\pm 1\}$ i.i.d., independent of the $w_r(0)$. Only the first layer is
trained: the $a_r$ are fixed at their initial values and gradient descent updates $W$ on the
squared loss
\begin{align}\label{eq:loss}
L(W) \;=\; \tfrac{1}{2} \norm{y - u(W)}^2,
\qquad u_i(W) := f_{W}(x_i),
\end{align}
with the update rule, for step size $\eta > 0$,
\begin{align}\label{eq:gd}
W(k+1) \;=\; W(k) - \eta \, \nabla_{W} L\!\left( W(k) \right),
\qquad k = 0, 1, 2, \ldots.
\end{align}
We abbreviate $u(k) := u(W(k))$ and write the residual as $y - u(k) \in \R^n$.

\begin{definition}[Empirical and population NTK Gram matrices]\label{def:gram}
For weights $W$, the \emph{empirical Gram matrix} $H(W) \in \R^{n \times n}$ has entries
\begin{align}\label{eq:emp-gram}
H_{ij}(W)
\;=\;
\frac{1}{m} \sum_{r=1}^{m} x_i^{\top} x_j \,
\indic{w_r^{\top} x_i \ge 0}\,\indic{w_r^{\top} x_j \ge 0}.
\end{align}
The \emph{population NTK Gram matrix} $\Hinf \in \R^{n \times n}$ has entries
\begin{align}\label{eq:pop-gram}
\Hinf_{ij}
\;=\;
\E_{w \sim \N(0, I_d)}\!\left[ x_i^{\top} x_j \,
\indic{w^{\top} x_i \ge 0}\,\indic{w^{\top} x_j \ge 0} \right]
\;=\;
\E\!\left[ H_{ij}(W(0)) \right].
\end{align}
We abbreviate $H(k) := H(W(k))$. Both matrices are symmetric positive semidefinite.
\end{definition}

\begin{assumption}[Spectral gap of the NTK]\label{ass:lambda0}
The population Gram matrix is strictly positive definite:
\begin{align*}
\lzero \;:=\; \lmin\!\left( \Hinf \right) \;>\; 0 .
\end{align*}
\end{assumption}

\begin{remark}\label{rem:lambda0-mild}
\Cref{ass:lambda0} is the \emph{single} spectral assumption of the analysis; every constant
below is tracked as a polynomial in $1/\lzero$. It is mild: $\lzero > 0$ holds whenever no two
inputs are parallel ($x_i \not\parallel x_j$ for $i \ne j$), which is generic for distinct
normalized data and is the condition identified by \cite{du2019gradient}. The matrix $\Hinf$
is exactly the Neural Tangent Kernel Gram matrix induced by the first-layer ReLU features at
infinite width.
\end{remark}

We record two standard external facts used repeatedly below.

\begin{fact}[Hoeffding's inequality, {\cite[Thm.~2.2.6]{vershynin2018high}}]\label{fac:hoeffding}
Let $Z_1, \dots, Z_m$ be independent random variables with $Z_r \in [\alpha_r, \beta_r]$
almost surely, and let $S = \sum_{r=1}^m Z_r$. Then for every $t > 0$,
\begin{align*}
\Pr\!\left[ \abs{S - \E S} \ge t \right]
\;\le\;
2 \exp\!\left( - \frac{2 t^2}{\sum_{r=1}^m (\beta_r - \alpha_r)^2} \right).
\end{align*}
\end{fact}

\begin{fact}[Weyl's inequality, {\cite[Cor.~III.2.6]{bhatia1997matrix}}]\label{fac:weyl}
Let $A, B \in \R^{n \times n}$ be symmetric. Then for every $i \in [n]$,
$\abs{\lambda_i(A) - \lambda_i(B)} \le \opnorm{A - B}$, where eigenvalues are sorted in
decreasing order. In particular $\lmin(A) \ge \lmin(B) - \opnorm{A - B}$.
\end{fact}
