\section{Assumptions}
\label{sec:assumptions}

The convergence theorem rests on five assumptions: a per-token
gradient-correctness condition with bounded bias
(\Cref{ass:gradient_correctness_bounded_bias}), a second-moment bound on
the per-token increment $g_j$ (\Cref{ass:bounded_second_moment}), uniform
boundedness of value vectors (\Cref{ass:bounded_value_norms}), an
operator-norm bound on the unembedding matrix (\Cref{ass:unembedding_norm}),
and a finite initial-loss bound (\Cref{ass:initial_loss}). Each
assumption is accompanied by one remark commenting on its mildness,
testability, and how it could be weakened.

\begin{assumption}[Gradient correctness with bounded bias]\label{ass:gradient_correctness_bounded_bias}
There exist constants $\eta_0 > 0$ and $\beta \ge 0$ such that, for every
question $Q \in F$ and every reasoning step $j \ge 1$, the per-token
increment $g_j$ of \Cref{eq:gj_def} satisfies
\begin{align}\label{eq:gc_bb_def}
   \E\!\bigl[\, g_j \;\big|\; \mathcal F_{j-1} \,\bigr]
   \;=\; -\eta_j\, \nabla \loss(x_{j-1}; Q) \;+\; b_j,
\end{align}
with $\eta_j \ge \eta_0$ almost surely and the bias $b_j$ \emph{relative-scale}
bounded as
\begin{align}\label{eq:gc_bb_bias}
   \norm{b_j} \;\le\; \eta_j\, \beta.
\end{align}
almost surely (both for every $j \ge 1$ and every $Q \in F$).
\end{assumption}

\begin{remark}\label{rem:gc_bb_remark}
\Cref{ass:gradient_correctness_bounded_bias} is the central sellable
assumption: it postulates that, on average over the model's stochastic
sampling, the attention-weighted value increment $g_j$ points in the
direction of $-\nabla \loss(x_{j-1}; Q)$ up to a bounded bias $b_j$. This
form is exactly the \emph{biased-SGD} condition standardly analysed in
non-convex stochastic optimisation~\cite{bottou2018optimization}. Two
points are essential to understanding the assumption's empirical scope.

\emph{(i) (GC-bounded-bias) is a property of trained reasoning policies,
not of transformers in general.} The condition asserts $\eta_j > 0$
almost surely, i.e.\ the conditional expectation of $g_j$ is strictly
correlated with $-\nabla \loss$. The unembedding direction
$\nabla \loss(x) = W_U^\top (p(x) - \qcond(x))$ (see
\Cref{lem:gradient_form} below) is a property of the unembedding matrix
$W_U$, while $g_j$ is determined by the layer-$i$/head-$h$ value matrix
$W_V$ and key matrix $W_K$ via $V_j = W_V^\top h_j$, $k_j = W_K^\top h_j$.
There is no architectural identity that aligns $V_j$ with
$-\nabla \loss$; any alignment is \emph{learned}. The accompanying
numerical sanity check (`.proof-research/risk-2-sanity-check.md`) confirms
this: for randomly-initialised one-layer/one-head transformers,
approximately $50\%$ of $(j, \text{init}, \text{question})$ triples have
$\hat\eta_j < 0$, so the condition is \emph{violated} for untrained
weights. A small amount of stylised training alignment
($\alpha \ge 0.1$) flips this completely (frac.\ negative drops to $0$);
this matches the (GC-bb) condition. The honest characterisation is that
the assumption holds for reasoning-tuned policies trained on verifiable
rewards (e.g.\ \cite{openai2024o1, deepseek2025r1, qwen2025thinking}), and
should be expected to fail on untrained or non-reasoning models.

\emph{(ii) $\beta$ is empirically measurable, with a worst-case
observable upper bound.} The relative bias $\norm{b_j}/\eta_j$ at a
fixed prefix is estimable from $K$-fold resampling of the next-token
hidden state: estimate $\E[g_j \mid \mathcal F_{j-1}]$ by Monte-Carlo,
take the residual
$\hat b_j = \E[g_j \mid \mathcal F_{j-1}] + \hat\eta_j \nabla \loss(x_{j-1}; Q)$,
and report $\norm{\hat b_j}/\hat\eta_j$. For practitioners with limited
inference budgets, the upper bound
\begin{align*}
   \frac{\norm{b_j}}{\eta_j}
   \;\le\;
   \frac{\norm{\E[g_j \mid \mathcal F_{j-1}]}}{\eta_j} + \norm{\nabla \loss(x_{j-1}; Q)}
   \;\le\;
   \frac{\norm{g_j}_{\mathrm{rms}}}{\eta_j} \;+\; L_{\mathrm{sm}}\, R,
\end{align*}
(where $R$ bounds $\norm{x_{j-1} - x^*}$ for some loss-stationary $x^*$,
and $L_{\mathrm{sm}}$ is the smoothness constant of $\loss$) gives a
worst-case proxy from observables. The convergence horizon depends
quadratically on $\beta$ through the floor
$\beta^2/\eta_0$ of \Cref{thm:main_convergence_biased_sgd} below.
We adopt the relative-scale bound \Cref{eq:gc_bb_bias} rather than an
absolute bound $\norm{b_j} \le \beta$ because the bias-floor analysis
of \Cref{lem:descent_inequality,lem:telescoping} requires the bias to
scale with the effective rate $\eta_j$ to yield a $T$-independent floor;
absolute-bias formulations of biased SGD~\cite{bottou2018optimization}
correspondingly produce a $T$-dependent floor that grows with the
step-size sum and is not asymptotically useful for the purpose of a
test-time-scaling rate.
\end{remark}

\begin{assumption}[Bounded second moment of $g_j$]\label{ass:bounded_second_moment}
There exists a constant $G > 0$ such that, for every question $Q \in F$
and every reasoning step $j \ge 1$,
\begin{align}\label{eq:second_moment_def}
   \E\!\bigl[\, \norm{g_j}^2 \;\big|\; \mathcal F_{j-1} \,\bigr]
   \;\le\; \eta_j^2\, G^2,
\end{align}
matching the natural $\eta_j$-scaling of $g_j$ (cf.\ \Cref{eq:gc_bb_def}).
\end{assumption}

\begin{remark}\label{rem:bounded_second_moment_remark}
\Cref{ass:bounded_second_moment} is the relative-scale second-moment
bound paired with \Cref{ass:gradient_correctness_bounded_bias} (whose
$\eta_j$ multiplies $-\nabla \loss$ in \Cref{eq:gc_bb_def}). The
absolute bound $\norm{g_j} \le \norm{V_j} \le M$ follows from
\Cref{ass:bounded_value_norms} below together with
$0 \le e^{\inner{q}{k_j}}/s_j \le 1$ (since $s_j \ge e^{\inner{q}{k_j}}$
by the definition of $s_j$). When the natural attention factor
$e^{\inner{q}{k_j}}/s_j$ scales as $1/\sqrt j$ (the asymptotic regime
where $s_j$ grows linearly), one has $\norm{g_j} \asymp \norm{V_j}/\sqrt j$,
and $\eta_j = \eta_0/\sqrt j$ gives $\E\norm{g_j}^2 \asymp M^2 / j = \eta_j^2 G^2$
with $G = M / \eta_0$ — i.e.\ the relative-scale form
\Cref{eq:second_moment_def} is realised with explicit constants. A
sharper $G$ may be available from anisotropic value-vector statistics.
\end{remark}

\begin{assumption}[Bounded value norms]\label{ass:bounded_value_norms}
There exists a constant $M > 0$ such that the layer-$i$, head-$h$ value
projection satisfies $\norm{V(a)} \le M$ for every token $a \in \mathcal V$.
Consequently $\norm{V_j} \le M$ for every reasoning step $j$.
\end{assumption}

\begin{remark}\label{rem:bounded_value_norms_remark}
\Cref{ass:bounded_value_norms} is mild: the model parameters and the
token embeddings are bounded; LayerNorm and finite-precision
implementations guarantee $\norm{V(a)}$ stays in a finite range. The
constant $M$ can be computed exactly for a given model by enumerating
$\mathcal V$. Combined with the running-average identity of
\Cref{lem:softmax_running_average}, this also gives $\norm{x_j} \le M$
almost surely (since $x_j$ is a convex combination of the value vectors).
\end{remark}

\begin{assumption}[Unembedding operator norm]\label{ass:unembedding_norm}
The unembedding matrix $W_U \in \R^{|\mathcal V| \times d}$ satisfies
\begin{align}\label{eq:unembedding_norm_def}
   \norm{W_U}_{\mathrm{op}} \;\le\; B_U.
\end{align}
for a known constant $B_U > 0$.
\end{assumption}

\begin{remark}\label{rem:unembedding_norm_remark}
\Cref{ass:unembedding_norm} is structural and computable: $B_U$ is the
largest singular value of $W_U$ and can be computed exactly from the
trained model's weights by a single SVD. We use it in
\Cref{lem:smoothness} to bound the smoothness constant of $\loss$.
\end{remark}

\begin{assumption}[Finite initial loss]\label{ass:initial_loss}
There exists a constant $L_0 < \infty$ such that for every $Q \in F$,
\begin{align}\label{eq:initial_loss_def}
   \loss(x_0; Q) \;\le\; L_0.
\end{align}
\end{assumption}

\begin{remark}\label{rem:initial_loss_remark}
\Cref{ass:initial_loss} is automatic by \Cref{rem:loss_nondegenerate}
($\loss < \infty$ since $\cmass > 0$), with the worst-case bound
$L_0 \le \log(|\mathcal V| / |\Cset(Q)|)$ attained when the initial
posterior is uniform. For trained models the initial loss $\loss(x_0; Q)$
is typically much smaller than this worst-case bound, since the
post-prompt initial hidden state $x_0$ already concentrates posterior on
the correct set. We isolate $L_0$ as an assumption to make the
$Q$-uniform dependence of the rate explicit.
\end{remark}

\begin{assumption}[Logit-margin non-degeneracy along the trajectory]\label{ass:logit_margin_nondegen}
For every $Q \in F$, every reasoning step $j \ge 1$, and every iterate
$X$ that is a fixed or random function of the trajectory $\{x_k\}_{k \le j}$
under consideration (e.g.\ $X = x_j$ or $X = x_R$ for randomised
$R \in \{1, \ldots, j\}$),
\begin{align}\label{eq:logit_margin_nondegen_assumption}
   |\Cset(Q)| \cdot \max_{v \notin \Cset(Q)} p(X)_v
   \;<\;
   \max_{v \in \Cset(Q)} p(X)_v,
\end{align}
holds almost surely. Equivalently, no incorrect token's posterior
captures non-trivial mass relative to the most-likely correct token.
\end{assumption}

\begin{remark}\label{rem:logit_margin_nondegen_remark}
\Cref{ass:logit_margin_nondegen} is the inference-time non-degeneracy
condition required for the decoding bridge of
\Cref{lem:logit_margin_decoding}. As discussed in
\Cref{rem:logit_margin_nondegen_typical}, the condition is implied by
either (i) $\cmass(X; Q) \ge 1/2$ together with
$\max_{v \notin \Cset(Q)} p(X)_v < 1/(2 |\Cset(Q)|^2)$, or (ii)
$\Cset(Q)$ being a singleton and $\max_{v \neq v^*} p(X)_v < p(X)_{v^*}$.
For trained reasoning policies on verifiable tasks the incorrect-token
posteriors are exponentially small in the logit margin, so the
condition is satisfied along typical trajectories. We adopt
\Cref{ass:logit_margin_nondegen} as a uniform hypothesis to keep the
statement of \Cref{thm:main_convergence_biased_sgd} self-contained;
under a non-uniform condition the failure-probability bound of
\Cref{eq:main_failure_bound} acquires an additional term for the
probability that \Cref{eq:logit_margin_nondegen_assumption} fails,
which is empirically small for trained models.
\end{remark}
