\section{Preliminaries}
\label{sec:prelim}

\paragraph{Notation.}
We write $\Pr[\cdot]$ for probability and $\E[\cdot]$ for expectation under a fixed
underlying probability space rich enough to support the random variables that follow.
Real intervals are written $[a, b]$ with $a \le b$. For a real-valued random variable
$X$ and $\lambda \in \R$, $\E[e^{\lambda X}]$ denotes the moment-generating function
(MGF) of $X$ at $\lambda$, which may be $+\infty$; under the boundedness hypotheses
we use it is always finite. We do not require $\E X = 0$ globally; centering is
stated where needed.

\paragraph{Setup.}
Throughout, let $n \ge 1$ be an integer and let $X_1, \ldots, X_n$ be
\emph{independent} real-valued random variables. For each $i \in \{1, \ldots, n\}$
we fix constants $a_i \le b_i$ with $X_i \in [a_i, b_i]$ almost surely and write
$\mu_i := \E X_i \in [a_i, b_i]$. Let
\begin{align*}
S_n \;:=\; \sum_{i=1}^n X_i,
\qquad
\mu \;:=\; \E S_n \;=\; \sum_{i=1}^n \mu_i,
\qquad
v \;:=\; \sum_{i=1}^n (b_i - a_i)^2.
\end{align*}
The centered summands are $Y_i := X_i - \mu_i$; they satisfy $\E Y_i = 0$ and
$Y_i \in [a_i - \mu_i,\, b_i - \mu_i]$ almost surely, an interval of the same
length $b_i - a_i$ as the original support of $X_i$. The independence of the $X_i$
descends to the $Y_i$, since each $Y_i$ is a deterministic function of $X_i$ alone.

\begin{fact}[Range invariance under centering]\label{fac:range-invariance}
If $X \in [a, b]$ almost surely with $\E X = \mu$, then $Y := X - \mu$ satisfies
$Y \in [a - \mu, b - \mu]$ almost surely, with $\E Y = 0$ and
$(b - \mu) - (a - \mu) = b - a$.
\end{fact}

\noindent
The fact is immediate from linearity of expectation and an additive shift of the
support; we record it explicitly because the equality of \emph{range lengths}
$(b - a)$ before and after centering is what allows the right-hand side of
Hoeffding's inequality to be stated in terms of the raw ranges $b_i - a_i$ rather
than the centered ranges.
