\section{Anchor-mass lower bound via the score margin}
\label{sec:anchor_mass}

The decomposition \Cref{lem:anchor_decomposition} reduces the convergence
of $x_j$ to $V^*$ to two questions: (i) how close are the value vectors
$V_k$ for $k \in \mathcal A^{\mathrm{traj}}_j$ to $V^*$ (answered by
\Cref{lem:anchor_accuracy_bound}), and (ii) how much softmax mass do
those anchor positions absorb (answered here). The conversion from the
score-margin assumption \Cref{ass:score_margin} to a softmax-mass lower
bound is a one-display exponential algebra step.

\begin{lemma}[Anchor mass lower bound]\label{lem:anchor_mass_lb}
Fix $Q \in F$. Under \Cref{ass:score_margin}, for every $j \ge 1$ and
every realisation of the trajectory with non-empty
$\mathcal A^{\mathrm{traj}}_j \subseteq \{1, \dots, j\}$,
\begin{align}\label{eq:anchor_mass_bound}
   \sum_{k \in \mathcal A^{\mathrm{traj}}_j} w_{j,k}
   \;\ge\;
   1 \;-\; \frac{\big|\{1, \dots, j\} \setminus \mathcal A^{\mathrm{traj}}_j\big|}
                {\big|\mathcal A^{\mathrm{traj}}_j\big|} \cdot e^{-\Delta},
\end{align}
where $w_{j,k}$ is the softmax weight from \Cref{def:softmax_attention}
and $\Delta > 0$ is the score-margin constant of \Cref{ass:score_margin}.
\end{lemma}

\begin{proof}
The proof proceeds in three steps: a pointwise bound on the
exponentiated score gap, a sum-of-non-anchors versus sum-of-anchors
ratio bound, and a softmax-normalisation conversion.

\noindent\textbf{Step 1: pointwise exponentiated-score bound.}
Fix any non-anchor position
$\ell \in \{1, \dots, j\} \setminus \mathcal A^{\mathrm{traj}}_j$ and any
anchor position $k^\star \in \mathcal A^{\mathrm{traj}}_j$. By
\Cref{ass:score_margin}, applied to the pair
$(a, a') = (a_{k^\star}, a_\ell)$,
$\inner{q}{k_{k^\star}} - \inner{q}{k_\ell} \ge \Delta$, so
\begin{align*}
   e^{\inner{q}{k_\ell}}
   \;\le\;
   e^{-\Delta}\, e^{\inner{q}{k_{k^\star}}}.
\end{align*}
Averaging this bound over $k^\star \in \mathcal A^{\mathrm{traj}}_j$
(the set is non-empty by hypothesis), we obtain
\begin{align}\label{eq:pointwise_non_anchor_bound}
   e^{\inner{q}{k_\ell}}
   \;\le\;
   \frac{e^{-\Delta}}{\big|\mathcal A^{\mathrm{traj}}_j\big|}
   \sum_{k \in \mathcal A^{\mathrm{traj}}_j} e^{\inner{q}{k_k}}.
\end{align}

\noindent\textbf{Step 2: sum-of-non-anchors versus sum-of-anchors.}
Summing \Cref{eq:pointwise_non_anchor_bound} over all
$\ell \in \{1, \dots, j\} \setminus \mathcal A^{\mathrm{traj}}_j$,
\begin{align*}
   \sum_{\ell \notin \mathcal A^{\mathrm{traj}}_j} e^{\inner{q}{k_\ell}}
   \;\le\; & ~
   \big|\{1, \dots, j\} \setminus \mathcal A^{\mathrm{traj}}_j\big| \cdot
   \frac{e^{-\Delta}}{\big|\mathcal A^{\mathrm{traj}}_j\big|}
   \sum_{k \in \mathcal A^{\mathrm{traj}}_j} e^{\inner{q}{k_k}} \\
   \;=\; & ~ r \cdot \sum_{k \in \mathcal A^{\mathrm{traj}}_j} e^{\inner{q}{k_k}},
\end{align*}
where the first step follows by summing
\Cref{eq:pointwise_non_anchor_bound} over the non-anchor index set, and
the last step introduces the shorthand
$r \coloneqq |\{1, \dots, j\} \setminus \mathcal A^{\mathrm{traj}}_j| / |\mathcal A^{\mathrm{traj}}_j| \cdot e^{-\Delta}$.

\noindent\textbf{Step 3: convert to softmax mass.}
Write $A \coloneqq \sum_{k \in \mathcal A^{\mathrm{traj}}_j} e^{\inner{q}{k_k}}$
and $B \coloneqq \sum_{\ell \notin \mathcal A^{\mathrm{traj}}_j} e^{\inner{q}{k_\ell}}$.
Step 2 reads $B \le r \cdot A$ and $s_j = A + B$. Therefore
\begin{align*}
   \sum_{k \in \mathcal A^{\mathrm{traj}}_j} w_{j,k}
   \;=\; & ~ \frac{A}{A + B} \\
   \;=\; & ~ \frac{1}{1 + B/A} \\
   \;\ge\; & ~ \frac{1}{1 + r} \\
   \;\ge\; & ~ 1 - r,
\end{align*}
where the first step follows from
$\sum_{k \in \mathcal A^{\mathrm{traj}}_j} w_{j,k} = A/s_j = A/(A+B)$
(by \Cref{def:softmax_attention}), the second step is algebraic
division, the third step follows from $B/A \le r$ (Step 2) and
monotonicity of $x \mapsto 1/(1+x)$, and the last step follows from the
elementary inequality $1/(1+r) \ge 1 - r$ valid for all $r \ge 0$ (since
$(1+r)(1-r) = 1 - r^2 \le 1$). Substituting back the definition of $r$
yields \Cref{eq:anchor_mass_bound}, as desired.
\end{proof}

\begin{remark}\label{rem:anchor_mass_lb_remark}
The bound \Cref{eq:anchor_mass_bound} is informative when the
anchor-count fraction $|\mathcal A^{\mathrm{traj}}_j|/j$ is not vanishing.
Combined with the high-probability lower bound
$|\mathcal A^{\mathrm{traj}}_T| \ge p_0 T / 2$ from
\Cref{lem:anchor_count_lb}, we obtain
$|\{1,\dots,T\}\setminus\mathcal A^{\mathrm{traj}}_T| \le T$ and
$|\mathcal A^{\mathrm{traj}}_T| \ge p_0 T / 2$, so on the event
$\Ecal_1(T, \delta_1)$,
\begin{align*}
   \sum_{k \in \mathcal A^{\mathrm{traj}}_T} w_{T,k}
   \;\ge\; 1 - \frac{T}{p_0 T / 2} \cdot e^{-\Delta}
   \;=\; 1 - \frac{2}{p_0} e^{-\Delta}.
\end{align*}
For this lower bound to exceed any pre-specified $1 - \eta$ with
$\eta \in (0, 1)$, it suffices that
$\Delta \ge \log\!\big(2/(p_0 \eta)\big)$ --- a condition on the trained
model's attention biases, not on $T$.
\end{remark}
